---
title: "Homework Three"
author: "Charlie"
date: "3/8/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r, include=FALSE}
library(tidyverse)
library(haven)
require(haven)
library(ggplot2)
nsch2019 <- read_sas("C://Users/charl/Desktop/Potential Datasets/nsch_2019_topical.sas7bdat")
```
## Summary of the Data and Relevant Variables
The dataset previewed is the National Survey of Children's Health 2019, administered by the U.S Census Bureau. It was designed to "provide national and state-level estimates on key indicators of the health and well-being of children, their families and their communities"^[Child and Adolescent Health Measurement Initiative (2020). “2019 National Survey of Children’s Health: Guide to Topics and Questions”. Data Resource Center for Child and Adolescent Health supported by the U.S. Department of Health and Human Services, Health Resources and Services
Administration (HRSA), Maternal and Child Health Bureau (MCHB). Retrieved [03/08/21] from [www.childhealthdata.org].]and measures the demographics of surveyed households, as well as specific survey measures for one randomly chosen child within the household ("selected child"). Depending on the age of the randomly chosen child, one of three survey forms was used- abbreviated within the dataset as T1, T2, or T3. 

I would like to examine some basic demographics of the children surveyed. This document will clean and break down the state of household residence (FIPSST) and the age (SC_AGE_YEARS), sex (SC_SEX), and race (SC_RACE_R) of that household's selected child. Each of these variables was collected during the "prescreening" process via an online form sent to potential participating households.

## Cleaning the Variables
The original formatting and coding of the original variables can be seen below (household ID (HHID) is included because it functions as the unique identifier for each observation, its "name" within the dataset)
```{r, include=TRUE}
#Preview the original variables
select(nsch2019, "HHID","FIPSST","SC_AGE_YEARS","SC_SEX","SC_RACE_R")
```
I decided not to separate the columns for household-wide information and child-specific information for several reasons- primarily because it would result in a massive amount of missing values within the dataframe. Further, the sampling method selected only one child per household, minimizing confusion. However, it is worth noting that there is information about multiple adults in the household within one observation, which might require further manipulation in order to make adult data clear.

My first step is to rename pedantic or tricky variables. Both SC_AGE_YEARS and SC_SEX are sufficiently descriptive but simple. However, FIPSST and SC_RACE_R are neither descriptive nor simple. For FIPSST, I intend to recode the data to use official postal codes in order to improve ggplot visualizations, so FIPSST (the state FIPS code) is no longer a useful variable name. For SC_RACE_R, institutional recodes and subgrouping appear to have complicated the variable names for race, but for my purposes, SC_RACE is much less confusing. 

```{r, include=TRUE}
#Rename problematic variables and preview new dataframe
nsch2019 <- rename(nsch2019, "STATE" = "FIPSST") %>%
rename("SC_RACE" = "SC_RACE_R")
select(nsch2019, "HHID","STATE","SC_AGE_YEARS","SC_SEX","SC_RACE")
```
My next concern is turning the values under the variables into something comprehensible. The values under SC_AGE_YEARS can stay numerical, as they just communicate the age of the selected child (unfortunately, it appears that any age under 1 is coded as 0, not specifying the months). SC_RACE, SC_SEX, and STATE, however, all require recoding in order to be useful for my intended analysis. 

SC_RACE, according to the codebook^[Child and Adolescent Health Measurement Initiative (CAHMI) (2021). 2019 National Survey of Children’s Health: Child and Family Health Measures, National Performance and Outcome Measures, and Subgroups, SAS Codebook, Data Resource Center for Child and Adolescent Health supported by the U.S. Department of Health and Human Services, Health Resources and Services Administration (HRSA), Maternal and Child Health Bureau (MCHB). Retrieved 2/22/2021 from childhealthdata.org], is coded as 

1 = White alone\
2 = Black or African American alone\
3 = American Indian or Alaska Native alone\
4 = Asian alone\
5 = Native Hawaiian and Other Pacific Islander alone\
7 = Two or More Races\

I will create a new variable, SC_RACE_CHAR, which spells out each racial category. There are no N/A values for SC_RACE, so there is no need to recode them.
```{r, include=TRUE}
#Recode SC_RACE according to the codebook as SC_RACE_CHR and print table 
#showing the old values and recoded values side by side
nsch2019 <- nsch2019 %>% mutate('SC_RACE_CHR' = recode(SC_RACE,
                                            "1" = "White",
                                            "2" = "Black",
                                            "3" = "AmerInd",
                                            "4" = "Asian",
                                            "5" = "PacIsl",
                                            "7" = "Biracial")) 
select(nsch2019, SC_RACE, SC_RACE_CHR)
```
SC_SEX will be recoded the same way. According to the codebook, SC_SEX is coded as 

1 = Male\
2 = Female\

Again, there are no N/A values for sex, so there is no need to specify an N/A recode. This appears to be because the surveyors "imputed" (inferred) certain demographics- there is a separate variable, SC_SEX_IF, which indicates whether or not the sex was imputed. This variable might be useful later on to filter "true" sex values from "guesses".
```{r, include=TRUE}
#Recode SC_SEX according to codebook as SC_SEX_CHR and print table
#showing the old values and recoded values side by side
nsch2019 <- nsch2019 %>% mutate('SC_SEX_CHR' = recode(SC_SEX,
                                            "1" = "M",
                                            "2" = "F")) 
select(nsch2019, SC_SEX, SC_SEX_CHR)
```
The final recode will be for STATE and will follow a similar format, just much (much) longer. I will be recoding based on the FIPS code to the two letter postal abbreviation ^[United States Department of Agriculture. (n.d.). State FIPS Codes. Natural Resources Conservation Service. Retrieved March 15, 2021, from https://www.nrcs.usda.gov/wps/portal/nrcs/detail/?cid=nrcs143_013696]. The way in which the survey was distributed necessarily included an FIPS code, so there are no N/A values and no need to recode for N/A.
```{r, include=TRUE}
#Recode STATE according to FIPS to POSTAL guide as STATE_POSTAL and print table
#showing the old values and recoded values side by side
nsch2019 <- nsch2019 %>% mutate('STATE_POSTAL' = recode(STATE,
                                            "01" = "AL",
                                            "02" = "AK",
                                            "04" = "AZ",
                                            "05" = "AR",
                                            "06" = "CA",
                                            "08" = "CO",
                                            "09" = "CT",
                                            "10" = "DE",
                                            "11" = "DC",
                                            "12" = "FL",
                                            "13" = "GA",
                                            "15" = "HI",
                                            "16" = "ID",
                                            "17" = "IL",
                                            "18" = "IN",
                                            "19" = "IA",
                                            "20" = "KS",
                                            "21" = "KY",
                                            "22" = "LA",
                                            "23" = "ME",
                                            "24" = "MD",
                                            "25" = "MA",
                                            "26" = "MI",
                                            "27" = "MN",
                                            "28" = "MS",
                                            "29" = "MO",
                                            "30" = "MT",
                                            "31" = "NE",
                                            "32" = "NV",
                                            "33" = "NH",
                                            "34" = "NJ",
                                            "35" = "NM",
                                            "36" = "NY",
                                            "37" = "NC",
                                            "38" = "ND",
                                            "39" = "OH",
                                            "40" = "OK",
                                            "41" = "OR",
                                            "42" = "PA",
                                            "44" = "RI",
                                            "45" = "SC",
                                            "46" = "SD",
                                            "47" = "TN",
                                            "48" = "TX",
                                            "49" = "UT",
                                            "50" = "VT",
                                            "51" = "VA",
                                            "53" = "WA",
                                            "54" = "WV",
                                            "55" = "WI",
                                            "56" = "WY")) 
select(nsch2019, STATE, STATE_POSTAL)
```
## Summary of New Variables
Through renaming and recoding, I now have the following variables to work with:

**SC_RACE_CHR-** An easier to understand description of the selected child's race\
**SC_AGE_YEARS-** Selected child's age in years (if over 11 months, otherwise 0)\
**STATE_POSTAL-** Selected child's *household's* state of residence at the moment of survey, displayed as the US Postal code\
**SC_SEX_CHR-** Selected child's sex at birth, displayed as M or F\
```{r, include=TRUE}
#Preview a dataframe with the relevant variables, sorted by household ID
nsch2019 %>%
  arrange("HHID") %>%
  select("HHID","SC_AGE_YEARS","STATE_POSTAL","SC_SEX_CHR","SC_RACE_CHR")
```

## Visualizing the New Variables
Now that I have variables which are easier to locate and understand, they can be used to create visualizations that reveal vital information about the demographics within the dataset. 

```{r, include=TRUE}
newbar <- ggplot(nsch2019, aes(STATE_POSTAL, fill=SC_RACE_CHR)) +
  geom_bar() + 
  labs(title= "NSCH State Participation", y="# of Respondents", x="State Postal Code", fill= "Race")
print(newbar)
```
Unfortunately, this graph is difficult to view within the Markdown document due to the many postal codes, but it is still evident that there is a fairly even count of participants from each state (and from DC, which indicates to me that the survey administrators likely wanted a certain sample size from each area). Adding SC_RACE_CHR as a fill allows me to further visualize the racial makeup of responses from each state- for example, we can see that Hawaii accounts for significantly fewer responses to the survey but has a larger proportion of Biracial and Asian respondents than any other state.

## Moving Forward
The demographic cleaning conducted above is not all that helpful alone (although it isn't entirely useless, as seen from the above graphic). However, it will help immensely in comparing demographic data to other, more meaningful behavioral data within the dataset. It adds potential for additional data manipulation- for example, STATE_POSTAL can be further simplified by organizing the postal codes into the Bureau of Economic Analysis' regional codes. It also creates important tools for visualization- for example, SC_RACE_CHR and SC_SEX can now be added to many different ggplots as a fill, to further understand demographic disparities in data. More advanced visualizations can focus on survey questions within a state- for example, among children in Georgia, which age group is most likely to have been to the dentist in the past year? 

My primary interest moving forward with this dataset is to explore racial and sex disparities in health data (as well as class disparities, but I am still struggling to grapple with this dataset's way of denoting income level), particularly between my home state (Georgia) and my current state of residence (Massachusetts). The above recoding is vital to my ability to conduct this analysis.